Climatic warming strengthens a positive feedback\\ between alpine shrubs and fire

### 2016-04-11 manuscript/manuscript.Rnw
Formatted manuscript for PNAS

\documentclass{nature}
\bibliographystyle{naturemag.bst}
\usepackage{lineno}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage[hidelinks]{hyperref}

\title{Climatic warming strengthens a positive feedback\\ between alpine shrubs and fire}
\author{James S. Camac$^{*,1,2,3}$, Richard J. Williams$^4$, Carl-Henrik Wahren$^3$, Ary A. Hoffmann$^5$ \\
\& Peter A. Vesk$^2$}

\begin{document}
\linenumbers
\maketitle

\begin{affiliations}
\item Department of Biological Sciences, Macquarie University, Sydney, NSW 2109, Australia.
\item The Centre of Excellence for Environmental Decisions, School of BioSciences, The University of Melbourne, Melbourne, Victoria 3010, Australia.
\item Research Centre for Applied Alpine Ecology, La Trobe University, Melbourne, Victoria 3086, Australia.
\item CSIRO Land and Water, Tropical Ecosystems Research Centre, PMB 44 Winnellie, Northern Territory 0821, Australia.
\item Bio21 Institute, School of BioSciences, The University of Melbourne, Victoria 3010, Australia.
\textasteriskcentered Email for correspondence: \texttt{james.camac@gmail.com}
\end{affiliations}

\pagebreak

@@

Accurately forecasting the effects of climatic warming on vegetation dynamics requires an understanding of the mechanisms by which climate and vegetation interact. Most forecasting models include the direct effects of climatic conditions on species distributions, but largely ignore biotic interactions and the type, frequency and severity of disturbances that are also likely to change\cite{Guisan:2005gz,Tucker:2012dr,Giannini:2013gi}. Because disturbance strongly influences recruitment opportunities, and thus, composition and structure of plant communities and biomes worldwide\cite{Bond:2005eg, Turner:2010wo}, it has the capacity to exacerbate or diminish how climate affects vegetation. Seedlings are the life stage most vulnerable to climate and disturbance, and the life stage which determines the long-term persistence of a species and its capacity to establish in new areas. Consequently, it is imperative to understand how seedlings respond to both changing climate and disturbance regimes\cite{Walck:2011dm, Briceno:2015iv}.

Field manipulative experiments have been widely used to investigate the effects of climate change on plant communities. These experiments have provided invaluable information on how climate change directly influences plant phenology\cite{Hoffmann:2010kf, Dorji:2013tt}, reproduction\cite{Klady:2011dw}, morphology\cite{Hudson:2009cn}, growth\cite{Hollister:2005vf}, floristic composition\cite{Elmendorf:2012kz} and biotic interactions\cite{Klanderud:2005uw}. However, because disturbances vary both spatially and temporally, most have focused on mature plant responses in relatively undisturbed vegetation\cite{Briceno:2015iv}. As such, few field climate experiments have been conducted on seedlings or vegetation in post-disturbance conditions (but see\cite{Munier:2010cg,Graae:2010ir,Camac:2015hja}). Consequently, little is known about how climate change influences seedling demographic rates\cite{Briceno:2015iv} and thus, vegetation recovery in post-disturbance conditions. We consider these issues in Australian alpine heathland.

@@

While paleoecological studies have indicated that such a feedback may exist\cite{Higuera:2008dw, Higuera:2009ga}, there is a paucity of information on multiple mechanisms that may strengthen, mitigate or break this feedback loop (Fig. \ref{fig1}a). For example, we have little information on whether fire will stimulate shrub seedling recruitment in alpine or tundra ecosystems, nor do we know how rates of seedling growth and survival will be affected under warmer, more exposed, post-fire conditions. Furthermore, we do not know how tussock grasslands (the other dominant community in these ecosystems) will respond to warmer post-fire conditions, whether grasses have competitive, facilitative or no effect on shrub seedling vital rates, or whether such effects are altered by warmer post-fire conditions.

@@

\subsection{Drivers of shrub seedling establishment.}
We first investigated how altitude, Topographic Wetness Index (TWI), adult density, fire and fire severity (as measured by post-fire twig diameters\textemdash see Supplementary Methods) influenced the density of \textit{Grevillea} and \textit{Asterolasia} seedlings (the two dominant shrubs of alpine open heathland). In 2011-12, we estimated seedling density for both species across 40 alpine sites and found that the abundances of \textit{Grevillea} (Fig. \ref{fig2}) and \textit{Asterolasia} (Supplementary Fig. S1) seedlings were strongly influenced by the occurrence of fire. Sites burnt in 2003 had seedling densities between \Sexpr{density_ratios(asttry_seedling_density_model,'Asterolasia')} and \Sexpr{density_ratios(greaus_seedling_density_model,'Grevillea')} times higher than unburnt sites. The mean seedling density of \textit{Grevillea}, was \Sexpr{density_means(greaus_seedling_density_model,'Grevillea', TRUE)}/m$^2$ at burnt sites and \Sexpr{density_means(greaus_seedling_density_model,'Grevillea', FALSE)}/m$^2$ at unburnt sites. \textit{Asterolasia} had similar mean densities: \Sexpr{density_means(asttry_seedling_density_model,'Asterolasia', TRUE)} and \Sexpr{density_means(asttry_seedling_density_model, 'Asterolasia', FALSE)} seedlings/m$^2$ at burnt and unburnt sites, respectively. For both species, seedling density was also higher in severely burnt sites (i.e. sites with larger post-fire twig diameters). As expected, pre-fire adult density positively influenced \textit{Grevillea} seedling density. For both shrub species, we detected no change in seedling density along a \Sexpr{round(max(complete_field_survey_data$altitude_m)) - round(min(complete_field_survey_data$altitude_m))} m altitudinal range (equivalent to a \Sexpr{round(0.77 * (200/100),1)}\textdegree C change in mean temperature\cite{Brown:1989vx}). The field warming experiment (see below) indicates that this lack of elevation response in \textit{Grevillea} is due to rates of recruitment (Supplementary Fig. S2) and mortality (see below) largely being insensitive to a \Sexpr{micro_clim(overall_microclimate_summary, 'CAmbient_Temp','mean','grand_mean')}\textdegree C change in temperature. We detected no strong effect of Topographic Wetness Index for either species.

\subsection{Experimental warming effects on shrub seedling growth.}

As shrub seedlings were more abundant in burnt vegetation, we investigated how warmer post-fire conditions affected seedling growth and mortality rates. Seedlings of \textit{Grevillea}, \textit{Asterolasia}, \textit{Phebalium} and \textit{Prostanthera} emerging after a wildfire were transplanted into experimentally burnt plots. These plots were either enclosed in Open Top Chambers (OTCs) whose temperature was \Sexpr{micro_clim(overall_microclimate_summary, 'CAmbient_Temp','mean','grand_mean')}\textdegree C warmer than control plots subjected to current ambient conditions. After \Sexpr{n_days(chamber_dates, FALSE)} days (\Sexpr{n_days(chamber_dates, TRUE)} growing season days) or \Sexpr{round(n_days(chamber_dates, FALSE)/365.25)} years' growth, mean seedling heights of the tall shrubs (\textit{Grevillea}, \textit{Phebalium} and \textit{Prostanthera}) growing in post-fire bare ground were greater in warmed plots relative to controls by \Sexpr{ht_diff(non_tussock_otc_growth_data,'Grevillea')}, \Sexpr{ht_diff(non_tussock_otc_growth_data,'Phebalium')} and \Sexpr{ht_diff(non_tussock_otc_growth_data,'Prostanthera')} cm, respectively (Fig. \ref{fig3}a). Warming increased heights of both \textit{Grevillea} and \textit{Prostanthera} seedlings in all years, while \textit{Phebalium} did not respond to the warming treatment until the second growing season. By contrast, seedlings of the prostrate shrub, \textit{Asterolasia}, showed no difference in growth rate between warmed and control plots in any year. For each species, we also observed similar treatment trends with stem diameter growth (Supplementary Fig. S3). Accounting for initial height and assuming logistic growth, the rates of change in mean annual predicted height of \textit{Grevillea}, \textit{Phebalium} and \textit{Prostanthera} were \Sexpr{growth_rate_ratio(non_tussock_otc_growth_model,'Grevillea')}, \Sexpr{growth_rate_ratio(non_tussock_otc_growth_model,'Phebalium')} and \Sexpr{growth_rate_ratio(non_tussock_otc_growth_model,'Prostanthera')} times that observed in control plots, respectively. According to this model, a 6 cm seedling (the mean initial height of seedlings used in this experiment) attains maximum height 39 years sooner \Sexpr{hmax(non_tussock_otc_growth_model,'Grevillea')} (\textit{Grevillea}) or \Sexpr{hmax(non_tussock_otc_growth_model,'Prostanthera')} years sooner (\textit{Phebalium} and \textit{Prostanthera}) when warmed by \Sexpr{micro_clim(overall_microclimate_summary, 'CAmbient_Temp','mean','grand_mean')}\textdegree C (Fig. \ref{fig3}b). \textit{Asterolasia} was predicted to reach its maximum height in approximately \Sexpr{round(hmax(non_tussock_otc_growth_model,'Asterolasia', FALSE)[2])} to \Sexpr{round(hmax(non_tussock_otc_growth_model,'Asterolasia', FALSE)[1])} years, irrespective of warming treatment.

@@

\subsection{Experimental warming on shrub seedling mortality.}

@@

\subsection{Effects of tussock grass proximity on shrub seedling growth and mortality rates.}

@@

\subsection{Experimental warming and rates of gap infilling by tussock grass.}

@@

\subsection{Strengthening of the warming-shrub-fire feedback.}

Here we quantified several unknown interactions between fire, shrub-grass relationships and climate to extend a conceptual model of alpine shrub dynamics under climate change (Fig. \ref{fig1}b). We showed how these interactions strengthen a hypothesized feedback loop that can rapidly increase shrub cover in alpine and tundra ecosystems\cite{Higuera:2008dw}. The combination of rising temperatures and more frequent or severe fire creates conditions that allow shrub seedlings to establish in greater densities, and for tall shrubs, double their growth rates and potentially increase their survival. These demographic effects will ultimately result in shrub thickening and expansion into grasslands. Coupled with field\cite{Williams:2006tl}, experimental\cite{Fraser:NHT5Kx_W} and paleoecological\cite{Higuera:2008dw} evidence, which indicate that shrubs are the most flammable component of alpine and tundra ecosystems, our results suggest that flammable fuel loads will accumulate faster and cover a larger proportion of the alpine and arctic landscapes under a warmer environment. This further strengthens the feedback\cite{Higuera:2008dw, Higuera:2009ga} by potentially increasing the frequency or severity of fires, which then creates further recruitment opportunities for shrubs with little demographic impact of neighbouring tussock grass.

@@

@@

\begin{methods}

We investigated shrub dynamics in open heathland because it is a common and highly flammable plant community in the Australian Alps above 1600 m a.s.l\cite{Williams:2006tl}. It is also an ecotone between closed heathland (\textgreater 70\% shrub cover) and tussock grassland consisting of \textit{Grevillea australis} shrubs interspersed among a \textit{Poa hiemata} sward. Aerial photography has revealed that the plant community has acted as an invasion front of shrub expansion into grassland and has itself experienced shrub thickening by closed heathland dominants such as \textit{Prostanthera cuneata}\cite{McDougall:2003gb}.

\subsection{Open Top Chamber Experiment.}

@@

\subsection{Seedling gradient study.}

@@

\subsection{Data analysis.}

In total we built 11 hierarchical models to examine factors influencing post-fire recovery of tussock grass and shrub seedling establishment, growth and mortality. For each model we used Bayesian inference and fitted models in R \Sexpr{paste(version$major, version$minor, sep='.')} using package \texttt{rstan} \Sexpr{packageVersion("rstan")}. Data and source code for reproducing analysis and figures are available at: \url{https://github.com/jscamac/Alpine_Shrub_Experiment}. Additional information about experimental design and analysis can be found in Supplementary Methods.
\end{methods}

\section*{Acknowledgements} 
This research was funded through Australian Research Council Linkage Grants, partnered through the Department of Sustainability and Environment, Parks Victoria, Commonwealth Scientific and Industrial Research Organisation (CSIRO). Funding was also received from the Centre of Excellence for Environmental Decisions (CEED) and Holsworth Research Committee. J.S.C was a recipient of an Australian Postgraduate Award. Monica Camac, Brad Farmilo, Monica Hursburgh, Paulius Kviecinskas, Karen Stott, Imogen Fraser, Sarah Bartlett, Paul Smart, Steven Rumbold, Lachlan Yourn and Chris Jones provided assistance with field measurements. William Morris and John Baumgartner provided coding advice and Warwick Papst provided field logistic advice. Daniel Falster and Richard Fitzjohn for developing software that enabled these analyses to be readily reproducible. Special thanks to David Bowman, Ana{\"i}s Gibert, John Morgan, Kimberley Millers, Michaela Plein and Inka Veltheim for their comments on this manuscript.

\section*{Author contributions}
J.C. conceived, designed and performed the experiments, field surveys and analysis; C-H.W, R.W, A.H and P.V. supervised the development of this work, aided in data collection and provided statistical advice. All authors contributed to the writing of this manuscript.

\section*{Competing financial interests} 
The authors declare that they have no competing financial interests.

\bibliography{refs.bib}

\section*{Figures}

\begin{figure}
\begin{minipage}{\textwidth}
\includegraphics{figures/fig1.pdf}
\caption{\textbf{Conceptual diagram illustrating positive warming-shrub-fire feedback loop.} (\textbf{a}) Hypothesised positive feedback loop between fire, climatic warming and landscape flammability. Solid blue lines = known mechansims; Dashed black lines = mechansims that may exacerbate or diminish this feedback loop but which we have a paucity of information on. (\textbf{b}) Feedback loop based on new evidence obtained from experimental and field surveys conducted in this study. Also includes an example of a potential, but unlikely, effect (red dotted line) that could break this feedback.}
\label{fig1}
\end{minipage}
\end{figure}

\begin{figure}
\begin{minipage}{\textwidth}
\includegraphics{figures/fig2.pdf}
\caption{\textbf{Mean \textit{Grevillea australis} seedling density along gradients of burning, fire severity (twig diameter), altitude, Topographic Wetness Index (TWI) and adult density.} (\textbf{a}) Centered and standardized model coefficients (on the log scale); and effects of: (\textbf{b}) burning, (\textbf{c}) fire severity, (\textbf{d}) altitude, (\textbf{e}) topographic wetness and (\textbf{f}) adult density, in areas burnt by the 2003 fires. All bars and shaded areas indicate 95\% Bayesian Credible Intervals.}
\label{fig2}
\end{minipage}
\end{figure}

\begin{figure}
\begin{minipage}{\textwidth}
\includegraphics{figures/fig3.pdf}
\caption{\textbf{Observed and projected growth trajectories of four dominant Australian alpine shrubs.} (\textbf{a}) Observed height growth: Thick lines with error bars represent mean ($\pm$ 95\% confidence intervals) observed heights at each May census. Thin lines represent individual growth trajectories. (\textbf{b}) Mean ($\pm$ 95\% Bayesian Credible Interval) projected growth trajectories. In all cases, red and orange lines = seedlings growing in warmed (OTC) conditions and blue and light blue lines = seedlings growing in control conditions. Projections were based on an logistic growth model using initial seedling size of 6 cm (the mean initial height observed in the OTC experiment) and mean maximum heights observed in long-unburnt sites (i.e. \Sexpr{hmax_values(hmax,'Asterolasia')}, \Sexpr{hmax_values(hmax,'Grevillea')}, \Sexpr{hmax_values(hmax,'Phebalium')} \& \Sexpr{hmax_values(hmax,'Prostanthera')} cm for \textit{Asterolasia}, \textit{Grevillea}, \textit{Phebalium}, \& \textit{Prostanthera}, respectively).}
\label{fig3}
\end{minipage}
\end{figure}

\begin{figure}
\begin{minipage}{\textwidth}
\includegraphics{figures/fig4.pdf}
\caption{\textbf{Mean ($\pm$ 95\% Bayesian Credible Interval) annual probability of death for each species growing in control (blue) and warmed (red) conditions.}}
\label{fig4}
\end{minipage}
\end{figure}
\pagebreak

\begin{figure}
\begin{minipage}{\textwidth}
\includegraphics{figures/fig5.pdf}
\caption{\textbf{Experimental warming and inter-tussock gap size effects on \textit{Grevillea australis} seedlings.} Top rows = Growth rate effects, Second row = Mortality effects. (\textbf{a} \& \textbf{c}) Centered and standardized model coefficients; (\textbf{b} \& \textbf{d}) growth and mortality rate response curves along an inter-tussock gap size gradient in warmed (red) and control (blue) conditions. All error bars and shading are 95\% Bayesian Credible Intervals.} 
\label{fig5}
\end{minipage}
\end{figure}
\pagebreak

\begin{figure}
\begin{minipage}{\textwidth}
\includegraphics{figures/fig6.pdf}
\caption{\textbf{Effects of warming on rates of gap infilling by tussock grass.} (\textbf{a}) Model coefficients and (\textbf{b}) projected temporal change in size for an average 10 cm inter-tussock gap in warmed (red) and unwarmed (blue) conditions. All error bars and shading are 95\% Bayesian Credible Intervals.} 
\label{fig6}
\end{minipage}
\end{figure}

\end{document}

>>>>

\documentclass[9pt,twocolumn,twoside, lineno]{pnas-new}
% Use the lineno option to display guide line numbers if required.
% Note that the use of elements such as single-column equations
% may affect the guide line number alignment.

\templatetype{pnasresearcharticle} % Choose template 
% {pnasresearcharticle} = Template for a two-column research article
% {pnasmathematics} = Template for a one-column mathematics article
% {pnasinvited} = Template for a PNAS invited submission

\title{Climatic warming strengthens a positive feedback\\ between alpine shrubs and fire}

% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\author[a,b,c,1]{James S Camac}
\author[d,2]{Richard J Williams} 
\author[c,3]{Carl-Henrik Wahren}
\author[e,4]{Ary A Hoffmann}
\author[b,5]{Peter A Vesk}

\affil[a]{Department of Biological Sciences, Macquarie University, Sydney, NSW 2109, Australia}
\affil[b]{The Centre of Excellence for Environmental Decisions, School of BioSciences, The University of Melbourne, Melbourne, Victoria 3010, Australia}
\affil[c]{Research Centre for Applied Alpine Ecology, La Trobe University, Melbourne, Victoria 3086, Australia}
\affil[d]{CSIRO Land and Water, Tropical Ecosystems Research Centre, PMB 44 Winnellie, Northern Territory 0821, Australia}
\affil[e]{Bio21 Institute, School of BioSciences, The University of Melbourne, Victoria 3010, Australia}

% Please give the surname of the lead author for the running footer
\leadauthor{Camac}

% Please add here a significance statement to explain the relevance of your work
\significancestatement{Temperatures are rapidly increasing in alpine and tundra ecosystems worldwide. A consequence of this temperature increase has been significant increases in shrub cover and a predicted increase in the frequency or severity of fires. Here we use experiemental and observational data to explore whether there are antagonistic or synergistic feedbacks between temperature, fire and shrub demographic rates. We find that increases in temperatures, coupled with more frequent or severe fire is likely to exacerbate the rate of shrub cover increases, and consquently, increase landscape flammability.}

% Please include corresponding author, author contribution and author declaration information
\authorcontributions{J.C. conceived, designed and performed the experiments, field surveys and analysis; C-H.W, R.W, A.H and P.V. supervised the development of this work, aided in data collection and provided statistical advice. All authors contributed to the writing of this manuscript.}
\authordeclaration{The authors declare that they have no competing financial interests.}
\correspondingauthor{\textsuperscript{1}To whom correspondence should be addressed. E-mail: james.camac@gmail.com}

% Keywords are not mandatory, but authors are strongly encouraged to provide them. If provided, please include two to five keywords, separated by the pipe symbol, e.g:
\keywords{Bayesian models $|$ Biotic interactions $|$ Growth $|$ Recruitment $|$ Mortality}

@@

\dates{This manuscript was compiled on \today}
\doi{\url{www.pnas.org/cgi/doi/10.1073/pnas.XXXXXXXXXX}}

\begin{document}
% Optional adjustment to line up main text (after abstract) of first page with line numbers, when using both lineno and twocolumn options.
% You should only change this length when you've finalised the article contents.
\verticaladjustment{-2pt}

\maketitle
\thispagestyle{firststyle}
\ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.
\dropcap{A}ccurately forecasting the effects of climatic warming on vegetation dynamics requires an understanding of the mechanisms by which climate and vegetation interact. Most forecasting models include the direct effects of climatic conditions on species distributions, but largely ignore biotic interactions and the type, frequency and severity of disturbances that are also likely to change\cite{Guisan:2005gz,Tucker:2012dr,Giannini:2013gi}. Because disturbance strongly influences recruitment opportunities, and thus, composition and structure of plant communities and biomes worldwide\cite{Bond:2005eg, Turner:2010wo}, it has the capacity to exacerbate or diminish how climate affects vegetation. Seedlings are the life stage most vulnerable to climate and disturbance, and the life stage which determines the long-term persistence of a species and its capacity to establish in new areas. Consequently, it is imperative to understand how seedlings respond to both changing climate and disturbance regimes\cite{Walck:2011dm, Briceno:2015iv}.

Field manipulative experiments have been widely used to investigate the effects of climate change on plant communities. These experiments have provided invaluable information on how climate change directly influences plant phenology\cite{Hoffmann:2010kf, Dorji:2013tt}, reproduction\cite{Klady:2011dw}, morphology\cite{Hudson:2009cn}, growth\cite{Hollister:2005vf}, floristic composition\cite{Elmendorf:2012kz} and biotic interactions\cite{Klanderud:2005uw}. However, because disturbances vary both spatially and temporally, most have focused on mature plant responses in relatively undisturbed vegetation\cite{Briceno:2015iv}. As such, few field climate experiments have been conducted on seedlings or vegetation in post-disturbance conditions (but see\cite{Munier:2010cg,Graae:2010ir,Camac:2015hja}). Consequently, little is known about how climate change influences seedling demographic rates\cite{Briceno:2015iv} and thus, vegetation recovery in post-disturbance conditions.

@@

\begin{figure}
\includegraphics{figures/fig1.pdf}
\caption{Conceptual diagram illustrating positive warming-shrub-fire feedback loop. (A) Hypothesised positive feedback loop between fire, climatic warming and landscape flammability. Solid blue lines = known mechansims; Dashed black lines = mechansims that may exacerbate or diminish this feedback loop but which we have a paucity of information on. (B) Feedback loop based on new evidence obtained from experimental and field surveys conducted in this study. Also includes an example of a potential, but unlikely, effect (red dotted line) that could break this feedback.}
\label{fig1}
\end{figure}

While paleoecological studies have indicated that such a feedback may exist\cite{Higuera:2008dw, Higuera:2009ga}, there is a paucity of information on multiple mechanisms that may strengthen, mitigate or break this feedback loop (Fig. \ref{fig1}A). For example, we have little information on whether fire will stimulate shrub seedling recruitment in alpine or tundra ecosystems, nor do we know how rates of seedling growth and survival will be affected under warmer, more exposed, post-fire conditions. Furthermore, we do not know how tussock grasslands (the other dominant community in these ecosystems) will respond to warmer post-fire conditions, whether grasses have competitive, facilitative or no effect on shrub seedling vital rates, or whether such effects are altered by warmer post-fire conditions.

@@

\section*{Results and Discussion}

\begin{figure}
\centering
\includegraphics{figures/fig2.pdf}
\caption{Mean \textit{Grevillea australis} seedling density along gradients of burning, fire severity (twig diameter), altitude, Topographic Wetness Index (TWI) and adult density. (A) Centered and standardized model coefficients (on the log scale); and effects of: (B) burning, (C) fire severity, (D) altitude, (E) topographic wetness and (F) adult density, in areas burnt by the 2003 fires. All bars and shaded areas indicate 95\% Bayesian Credible Intervals.}
\label{fig2}
\end{figure}

\begin{figure*}
\centering
\includegraphics{figures/fig3.pdf}
\caption{Observed and projected growth trajectories of four dominant Australian alpine shrubs. (A) Observed height growth: Thick lines with error bars represent mean ($\pm$ 95\% confidence intervals) observed heights at each May census. Thin lines represent individual growth trajectories. (B) Mean ($\pm$ 95\% Bayesian Credible Interval) projected growth trajectories. In all cases, red and orange lines = seedlings growing in warmed (OTC) conditions and blue and light blue lines = seedlings growing in control conditions. Projections were based on an logistic growth model using initial seedling size of 6 cm (the mean initial height observed in the OTC experiment) and mean maximum heights observed in long-unburnt sites (i.e. \Sexpr{hmax_values(hmax,'Asterolasia')}, \Sexpr{hmax_values(hmax,'Grevillea')}, \Sexpr{hmax_values(hmax,'Phebalium')} \& \Sexpr{hmax_values(hmax,'Prostanthera')} cm for \textit{Asterolasia}, \textit{Grevillea}, \textit{Phebalium}, \& \textit{Prostanthera}, respectively).}
\label{fig3}
\end{figure*}

\subsection*{Drivers of shrub seedling establishment}
We first investigated how altitude, Topographic Wetness Index (TWI), adult density, fire and fire severity (as measured by post-fire twig diameters\textemdash see SI Materials and Methods) influenced the density of \textit{Grevillea} and \textit{Asterolasia} seedlings (the two dominant shrubs of alpine open heathland). In 2011-12, we estimated seedling density for both species across 40 alpine sites and found that the abundances of \textit{Grevillea} (Fig. \ref{fig2}) and \textit{Asterolasia} (Supplementary Fig. S1) seedlings were strongly influenced by the occurrence of fire. Sites burnt in 2003 had seedling densities between \Sexpr{density_ratios(asttry_seedling_density_model,'Asterolasia')} and \Sexpr{density_ratios(greaus_seedling_density_model,'Grevillea')} times higher than unburnt sites. The mean seedling density of \textit{Grevillea}, was \Sexpr{density_means(greaus_seedling_density_model,'Grevillea', TRUE)}/m$^2$ at burnt sites and \Sexpr{density_means(greaus_seedling_density_model,'Grevillea', FALSE)}/m$^2$ at unburnt sites. \textit{Asterolasia} had similar mean densities: \Sexpr{density_means(asttry_seedling_density_model,'Asterolasia', TRUE)} and \Sexpr{density_means(asttry_seedling_density_model, 'Asterolasia', FALSE)} seedlings/m$^2$ at burnt and unburnt sites, respectively. For both species, seedling density was also higher in severely burnt sites (i.e. sites with larger post-fire twig diameters). As expected, pre-fire adult density positively influenced \textit{Grevillea} seedling density. For both shrub species, we detected no change in seedling density along a \Sexpr{round(max(complete_field_survey_data$altitude_m)) - round(min(complete_field_survey_data$altitude_m))} m altitudinal range (equivalent to a \Sexpr{round(0.77 * (200/100),1)}\textdegree C change in mean temperature\cite{Brown:1989vx}). The field warming experiment (see below) indicates that this lack of elevation response in \textit{Grevillea} is due to rates of recruitment (Supplementary Fig. S2) and mortality (see below) largely being insensitive to a \Sexpr{micro_clim(overall_microclimate_summary, 'CAmbient_Temp','mean','grand_mean')}\textdegree C change in temperature. We detected no strong effect of Topographic Wetness Index for either species.

\subsection*{Experimental warming effects on shrub seedling growth}

As shrub seedlings were more abundant in burnt vegetation, we investigated how warmer post-fire conditions affected seedling growth and mortality rates. Seedlings of \textit{Grevillea}, \textit{Asterolasia}, \textit{Phebalium} and \textit{Prostanthera} emerging after a wildfire were transplanted into experimentally burnt plots. These plots were either enclosed in Open Top Chambers (OTCs) whose temperature was \Sexpr{micro_clim(overall_microclimate_summary, 'CAmbient_Temp','mean','grand_mean')}\textdegree C warmer than control plots subjected to current ambient conditions. After \Sexpr{n_days(chamber_dates, FALSE)} days (\Sexpr{n_days(chamber_dates, TRUE)} growing season days) or \Sexpr{round(n_days(chamber_dates, FALSE)/365.25)} years' growth, mean seedling heights of the tall shrubs (\textit{Grevillea}, \textit{Phebalium} and \textit{Prostanthera}) growing in post-fire bare ground were greater in warmed plots relative to controls by \Sexpr{ht_diff(non_tussock_otc_growth_data,'Grevillea')}, \Sexpr{ht_diff(non_tussock_otc_growth_data,'Phebalium')} and \Sexpr{ht_diff(non_tussock_otc_growth_data,'Prostanthera')} cm, respectively (Fig. \ref{fig3}A). Warming increased heights of both \textit{Grevillea} and \textit{Prostanthera} seedlings in all years, while \textit{Phebalium} did not respond to the warming treatment until the second growing season. By contrast, seedlings of the prostrate shrub, \textit{Asterolasia}, showed no difference in growth rate between warmed and control plots in any year. For each species, we also observed similar treatment trends with stem diameter growth (Supplementary Fig. S3). Accounting for initial height and assuming logistic growth, the rates of change in mean annual predicted height of \textit{Grevillea}, \textit{Phebalium} and \textit{Prostanthera} were \Sexpr{growth_rate_ratio(non_tussock_otc_growth_model,'Grevillea')}, \Sexpr{growth_rate_ratio(non_tussock_otc_growth_model,'Phebalium')} and \Sexpr{growth_rate_ratio(non_tussock_otc_growth_model,'Prostanthera')} times that observed in control plots, respectively. According to this model, a 6 cm seedling (the mean initial height of seedlings used in this experiment) attains maximum height 39 years sooner \Sexpr{hmax(non_tussock_otc_growth_model,'Grevillea')} (\textit{Grevillea}) or \Sexpr{hmax(non_tussock_otc_growth_model,'Prostanthera')} years sooner (\textit{Phebalium} and \textit{Prostanthera}) when warmed by \Sexpr{micro_clim(overall_microclimate_summary, 'CAmbient_Temp','mean','grand_mean')}\textdegree C (Fig. \ref{fig3}B). \textit{Asterolasia} was predicted to reach its maximum height in approximately \Sexpr{round(hmax(non_tussock_otc_growth_model,'Asterolasia', FALSE)[2])} to \Sexpr{round(hmax(non_tussock_otc_growth_model,'Asterolasia', FALSE)[1])} years, irrespective of warming treatment.

@@

\subsection*{Experimental warming on shrub seedling mortality}

@@

\subsection*{Effects of tussock grass proximity on shrub seedling growth and mortality rates}

@@

\subsection*{Experimental warming and rates of gap infilling by tussock grass}

@@

\subsection*{Strengthening of the warming-shrub-fire feedback}

Here we quantified several unknown interactions between fire, shrub-grass relationships and climate to extend a conceptual model of alpine shrub dynamics under climate change (Fig. \ref{fig1}B). We showed how these interactions strengthen a hypothesized feedback loop that can rapidly increase shrub cover in alpine and tundra ecosystems\cite{Higuera:2008dw}. The combination of rising temperatures and more frequent or severe fire creates conditions that allow shrub seedlings to establish in greater densities, and for tall shrubs, double their growth rates and potentially increase their survival. These demographic effects will ultimately result in shrub thickening and expansion into grasslands. Coupled with field\cite{Williams:2006tl}, experimental\cite{Fraser:NHT5Kx_W} and paleoecological\cite{Higuera:2008dw} evidence, which indicate that shrubs are the most flammable component of alpine and tundra ecosystems, our results suggest that flammable fuel loads will accumulate faster and cover a larger proportion of the alpine and arctic landscapes under a warmer environment. This further strengthens the feedback\cite{Higuera:2008dw, Higuera:2009ga} by potentially increasing the frequency or severity of fires, which then creates further recruitment opportunities for shrubs with little demographic impact of neighbouring tussock grass.

\begin{figure}
\centering
\includegraphics{figures/fig4.pdf}
\caption{Mean ($\pm$ 95\% Bayesian Credible Interval) annual probability of death for each species growing in control (blue) and warmed (red) conditions.}
\label{fig4}
\end{figure}

\begin{figure}
\centering
\includegraphics{figures/fig5.pdf}
\caption{Experimental warming and inter-tussock gap size effects on \textit{Grevillea australis} seedlings. Top rows = Growth rate effects, Second row = Mortality effects. (A \& C) Centered and standardized model coefficients; (B \& D) growth and mortality rate response curves along an inter-tussock gap size gradient in warmed (red) and control (blue) conditions. All error bars and shading are 95\% Bayesian Credible Intervals.} 
\label{fig5}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{figures/fig6.pdf}
\caption{Effects of warming on rates of gap infilling by tussock grass. (A) Model coefficients and (B) projected temporal change in size for an average 10 cm inter-tussock gap in warmed (red) and unwarmed (blue) conditions. All error bars and shading are 95\% Bayesian Credible Intervals.} 
\label{fig6}
\end{figure}

@@

\matmethods{
We investigated shrub dynamics in open heathland because it is a common and highly flammable plant community in the Australian Alps above 1600 m a.s.l\cite{Williams:2006tl}. It is also an ecotone between closed heathland (\textgreater 70\% shrub cover) and tussock grassland consisting of \textit{Grevillea australis} shrubs interspersed among a \textit{Poa hiemata} sward. Aerial photography has revealed that the plant community has acted as an invasion front of shrub expansion into grassland and has itself experienced shrub thickening by closed heathland dominants such as \textit{Prostanthera cuneata}\cite{McDougall:2003gb}.

\subsection*{Open Top Chamber Experiment}

@@

\subsection*{Seedling gradient study}

@@

\subsection*{Data analysis}

In total we built 11 hierarchical models to examine factors influencing post-fire recovery of tussock grass and shrub seedling establishment, growth and mortality. For each model we used Bayesian inference and fitted models in R \Sexpr{paste(version$major, version$minor, sep='.')} using package \texttt{rstan} \Sexpr{packageVersion("rstan")}. Data and source code for reproducing analysis and figures are available at: \url{https://github.com/jscamac/Alpine_Shrub_Experiment}. Additional information about experimental design and analysis can be found in SI Materials and Methods.
\\
\\
}

\showmatmethods % Display the Materials and Methods section

\acknow{
This research was funded through Australian Research Council Linkage Grants, partnered through the Department of Sustainability and Environment, Parks Victoria, Commonwealth Scientific and Industrial Research Organisation (CSIRO). Funding was also received from the Centre of Excellence for Environmental Decisions (CEED) and Holsworth Research Committee. J.S.C was a recipient of an Australian Postgraduate Award. Monica Camac, Brad Farmilo, Monica Hursburgh, Paulius Kviecinskas, Karen Stott, Imogen Fraser, Sarah Bartlett, Paul Smart, Steven Rumbold, Lachlan Yourn and Chris Jones provided assistance with field measurements. William Morris and John Baumgartner provided coding advice and Warwick Papst provided field logistic advice. Daniel Falster and Richard Fitzjohn for developing software that enabled these analyses to be readily reproducible. Special thanks to David Bowman, Ana{\"i}s Gibert, John Morgan, Kimberley Millers, Michaela Plein and Inka Veltheim for their comments on this manuscript.}

\showacknow % Display the acknowledgments section

% \pnasbreak splits and balances the columns before the references.
% If you see unexpected formatting errors, try commenting out this line
% as it can run into problems with floats and footnotes on the final page.
%\pnasbreak

% Bibliography
\bibliography{refs.bib}
%\pnasbreak
\end{document}
